---
title: "02b Control File — Complex Panels"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article covers control file setup for panels that are more complex than a
straightforward single-cell-type staining with one unstained sample. Specifically:

1. **Mixed bead and cell controls** — different universal negatives for each type
2. **Viability markers** — how the `is.viability` column affects the legacy
   gating pipeline (not the automated pipeline)
3. **Fluorophores not in the AutoSpectral database** — how to handle them and
   how to fill in their peak channel manually
4. **Fluorophore name disambiguation** when you have both beads and cells for
   the same target (not recommended, but possible)

This builds on the basics in
[02 Creating the Control File](https://drcytometer.github.io/AutoSpectral/articles/02_Control_File_example.html).
The automated pipeline (`get.spectra.automated()`) is used throughout.

```{r setup, eval=FALSE}
library(AutoSpectral)
asp <- get.autospectral.param(cytometer = "aurora")
control.dir <- "~/my_experiment/controls"
create.control.file(control.dir, asp)
control.file <- "fcs_control_file.csv"
```

# Mixed bead and cell controls

Many panels use compensation beads for dim or rare markers and cell controls
for bright or common ones. Both types can coexist in a single control file.

The `control.type` column tells AutoSpectral how to treat each sample. Set each
row to `"cells"` or `"beads"` (lower case). In the automated pipeline:

- **Cells** are processed with AF orthogonalisation and cosine-similarity
  filtering against the universal negative's AF vector. Scatter-matched
  AF subtraction uses the nearest events from the paired unstained cell sample.
- **Beads** are assumed to have homogeneous, low-AF backgrounds. Cosine
  filtering against an AF vector is skipped. The "universal" negative for a bead
  control should be an unstained bead sample. If your beads have separate negative
  and positive beads, use the unstained positive beads as the universal negative.

`check.control.file()` enforces that the `control.type` of a sample and its
`universal.negative` must match — a cell control paired with a bead negative
(or vice versa) is flagged as an error.

## Naming fluorophores when you have both beads and cells

`check.control.file()` requires every `fluorophore` value to be unique. If you
have both a bead and a cell single-stained control for the same fluorophore,
give them distinct names. For example:

| filename           | fluorophore | control.type | universal.negative      |
|--------------------|-------------|--------------|--------------------------|
| PE_cells.fcs       | PE          | cells        | Unstained_cells.fcs     |
| PE_beads.fcs       | PE beads    | beads        | Unstained_beads.fcs     |
| Unstained_cells.fcs| AF          | cells        | Unstained_cells.fcs     |
| Unstained_beads.fcs| Negative    | beads        | Unstained_beads.fcs     |

The name in the `fluorophore` column becomes the parameter name in the unmixed
FCS file, so pick whichever version you want to appear in your analysis
software. You will then run `get.spectra.automated()` once and receive spectra
for both `PE` and `PE beads`. Choose one for unmixing — typically whichever
produces the cleaner, more accurate spectrum (check the QC ribbon plots in
`figure_spectral_ribbon/`).

If you want to unmix using only one, subset the spectra matrix before passing
it to `unmix.fcs()`:

```{r, eval=FALSE}
# keep only the cell-based PE spectrum
spectra.subset <- spectra[ rownames(spectra) != "PE beads", ]
```

## Multiple unstained samples

When you have more than one unstained sample (one for cells, one for beads),
both must appear in the `filename` column. Set `fluorophore` to `AF` for the
unstained cell sample and to `Negative` for the unstained bead sample.
`check.control.file()` enforces that every value in `universal.negative` must
itself appear as a row in `filename`, and that it must be labelled `AF` or
`Negative`.

```{r, eval=FALSE}
issues <- check.control.file(control.dir, control.file, asp)
```

# Viability markers

A viability dye labels dead cells selectively. Dead cells sit to the left of
the live population on FSC. This only matters for the **legacy pipeline**
(`define.flow.control()`), which uses the `is.viability` column to widen the
FSC gate leftward to capture dead cells. The column is ignored by
`get.spectra.automated()`.

If you are using the automated pipeline, you do not need to fill in
`is.viability`. The automated extraction selects high-expressing events based
on the peak channel signal and cosine filtering rather than scatter gating, so
it works correctly for viability dyes without any special configuration.

For completeness, the legacy approach is:

```
| filename           | fluorophore | is.viability |
|--------------------|-------------|--------------|
| LiveDead_cells.fcs | eFluor780   | TRUE         |
```

For the v1.5.0 landmark gating system, assign the viability control to a
dedicated gate group by setting `gate.name` to something like `"dead"` and
`gate.define = TRUE`. See the
[Full Workflow](https://drcytometer.github.io/AutoSpectral/articles/01_Full_AutoSpectral_Workflow.html)
and [Advanced Gating](https://drcytometer.github.io/AutoSpectral/articles/08_Advanced_Gating.html)
articles for details.

# Fluorophores not in the database

If AutoSpectral cannot match a file name to a known fluorophore, you will see
`No Match` in the `fluorophore` column. This is a validation error —
`check.control.file()` will not let you proceed until every `No Match` is
replaced.

**Step 1 — fill in `fluorophore`.**
Write any name you like. This name will appear in the unmixed FCS file and in
all AutoSpectral plots, so choose carefully. The name does not need to match
any database entry for the pipeline to run.

**Step 2 — fill in `channel`.**
This is the expected peak detection channel for the fluorophore on your
cytometer. You can find it using:

- [Cytek Cloud](https://cloud.cytekbio.com/) (for Cytek instruments)
- [BD Spectrum Viewer](https://www.bdbiosciences.com/en-us/resources/bd-spectrum-viewer) (for BD instruments)
- [BioLegend Spectra Analyzer](https://www.biolegend.com/en-gb/spectra-analyzer)
- [FluoroFinder](https://fluorofinder.com/) — note that FluoroFinder channel
  names do not always match the actual names in the FCS file, so treat this as
  approximate and verify against the
  [cytometer database](https://docs.google.com/spreadsheets/d/1wj7QPkgpsuPNeVKyt-WWdBu5R48aZTgEbH8-_bpKeBY/edit?usp=sharing).

In the **automated pipeline**, the `channel` column is optional — if you leave
it blank, `check.control.file()` will issue a warning and `get.spectra.automated()`
will derive the peak channel empirically from the data. However, providing the
correct channel enables the QC step to verify the extraction against the
expected peak, which is strongly recommended. An unrecognised fluorophore will
not have a spectral reference library entry, so the cosine-similarity QC check
against the reference library will be skipped for that fluorophore; only the
peak-signal check will apply.

**Step 3 (optional but encouraged) — add to the database.**
If you add your fluorophore (with synonyms and peak channel entries for each
cytometer) to the
[fluorophore database](https://docs.google.com/spreadsheets/d/14j4lAQ6dkjDBKMborDv_MkSptyNBqZiBsq5jNNSCoiQ/edit?usp=sharing),
it will be recognised automatically by future versions of AutoSpectral. Changes
are incorporated with the next package update.

# Putting it together — example control file

Here is what a completed control file might look like for a panel with bead and
cell controls, an unstained bead sample, a viability dye, and one fluorophore
not in the database:

| filename               | fluorophore  | marker   | channel   | control.type | universal.negative      |
|------------------------|--------------|----------|-----------|--------------|--------------------------|
| BV421_cells.fcs        | BV421        | CD4      | V4-A      | cells        | Unstained_cells.fcs     |
| PE_beads.fcs           | PE           | CD14     | YG3-A     | beads        | Unstained_beads.fcs     |
| eFluor780_cells.fcs    | eFluor780    | Viability|           | cells        | Unstained_cells.fcs     |
| NewDye660_cells.fcs    | NewDye660    | CD8      | R2-A      | cells        | Unstained_cells.fcs     |
| Unstained_cells.fcs    | AF           |          |           | cells        | Unstained_cells.fcs     |
| Unstained_beads.fcs    | Negative     |          |           | beads        | Unstained_beads.fcs     |

Notes on this example:

- `eFluor780` has no `channel` entry — the automated pipeline will derive the
  peak empirically and warn about the missing channel. In practice it is best
  to fill this in.
- `NewDye660` is a hypothetical fluorophore not in the database. Its
  `channel` (`R2-A`) has been filled in manually.
- Both unstained samples are present, with matching `control.type` values for
  each paired control.

# Running the check

```{r, eval=FALSE}
issues <- check.control.file(control.dir, control.file, asp)
```

The check is run in automated mode (`legacy = FALSE` by default), so gating
columns are not validated. Pay attention to:

| Severity | What to look for |
|----------|-----------------|
| **Error** | `No Match` in fluorophore; duplicate fluorophore names; `control.type` mismatch between sample and its universal negative; universal negative filename not found; `AF` control absent or set to `beads` |
| **Warning** | Missing `channel` for non-AF/Negative rows; event count < 5,000; cytometer name mismatch |

Fix all errors before proceeding. When the check passes with a green message,
run spectral extraction:

```{r, eval=FALSE}
spectra <- get.spectra.automated(
  control.dir      = control.dir,
  control.def.file = control.file,
  asp              = asp
)
```

Check the outputs in `figure_spectral_ribbon/` to verify that spectra for both
bead and cell controls look reasonable, then select the best version of each
for unmixing.
