---
title: "16 Development"
output: html_document
date: "2026-03-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here's some stuff that's in progress. These changes will show up first in the development `dev` branch, which you are welcome to install and test.

If you would like to contribute to AutoSpectral, these are key areas I have identified where it has shortcomings or needs improvement. Suggestions for strategies, or better, code suggestions, would be appreciated.

* Unmixing with controls from different days or different voltages
 + AutoSpectral does not currently support this, although it may work reasonably well anyway since these modern spectral cytometers are fairly well calibrated for PMT/APD linearity.
 + In most cases, AutoSpectral does not have access to the information needed to normalize the files between different voltage settings. For Cytek instruments, this would be QC information and the initial linearity response curve for the APDs. For FACSDiscover systems, this is likely the QSPE values, which are in the FCS files.
 + I'd need to sit down and work out the normalization functions. This hasn't been a top priority.

* More flexibility in negative/unstained samples.

* Allow multiple controls per fluorophore.

* Fix the issue causing discontinuities.
  + Progress has been made on this in v1.0.0 via the new, sped-up optimization strategy. The "slow" method should now be fast enough to fix most of the issue.
  + Additional improvements include better decontamination of AF from the spectral variants, which was causing cells to be pushed further away from the threshold for optimization, and a rduction in low-level noise introduced into the spectral variants due to fluctuations in electronic noise or autofluorescence. Any signal in the variant spectrum in a channel where the optimized single spectrum has less than 5% of the peak detector signal is now regularized towards the optimized spectrum. This focuses the variation on the "peaks", which is what causes the unmixing errors and spread.
  + This should be eliminated entirely with the next round of updates, tentatively 1.6.0.
  
* Better correction of unmixing errors
  + There are several strategies I'm investigating to handling the cases where the multi-color samples contain obvious errors outside the range of variation seen in the single-color controls. This is not stuff I'm going to put online, so get in touch if you'd like to work on this. 
  + This should be improved further with the next round of updates, tentatively 1.6.0

To install the `dev` branch:

```{r dev branch, eval=FALSE}
devtools::install_github("DrCytometer/AutoSpectral@dev")
```

To replace this with the (hopefully) stable version, run this:

```{r master branch, eval=FALSE}
remove.packages("AutoSpectral")
devtools::install_github("DrCytometer/AutoSpectral")
```



