---
title: "16 Development"
output: html_document
date: "2026-03-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here's some stuff that's in progress. These changes will show up first in the development `dev` branch, which you are welcome to install and test.

If you would like to contribute to AutoSpectral, these are key areas I have identified where it has shortcomings or needs improvement. Suggestions for strategies, or better, code suggestions, would be appreciated.

* Unmixing with controls from different days or different voltages
 + AutoSpectral does not currently support this, although it may work reasonably well anyway since these modern spectral cytometers are fairly well calibrated for PMT/APD linearity.
 + In most cases, AutoSpectral does not have access to the information needed to normalize the files between different voltage settings. For Cytek instruments, this would be QC information and the initial linearity response curve for the APDs. For FACSDiscover systems, this is likely the QSPE values, which are in the FCS files.
 + I'd need to sit down and work out the normalization functions. This hasn't been a top priority.

* More flexibility in negative/unstained samples.

* Allow multiple controls per fluorophore.

* Fix the issue causing discontinuities.
  + Progress has been made on this in v1.0.0 via the new, sped-up optimization strategy. The "slow" method should now be fast enough to fix most of the issue.
  + Additional improvements include better decontamination of AF from the spectral variants, which was causing cells to be pushed further away from the threshold for optimization, and a rduction in low-level noise introduced into the spectral variants due to fluctuations in electronic noise or autofluorescence. Any signal in the variant spectrum in a channel where the optimized single spectrum has less than 5% of the peak detector signal is now regularized towards the optimized spectrum. This focuses the variation on the "peaks", which is what causes the unmixing errors and spread.
  + This should be eliminated entirely with the next round of updates, tentatively 1.6.0.
  
* Better correction of unmixing errors
  + There are several strategies I'm investigating to handling the cases where the multi-color samples contain obvious errors outside the range of variation seen in the single-color controls. This is not stuff I'm going to put online, so get in touch if you'd like to work on this. 
  + This should be improved further with the next round of updates, tentatively 1.6.0

## Benchmarking new AF assignment strategies

If you are developing a new autofluorescence (AF) assignment or fitting
function and want to compare it against the existing approaches, AutoSpectral
provides two benchmarking functions: `test.af.accuracy()` and
`benchmark.af.spectra.size()`.

### What they measure

Both functions evaluate AF assignment quality using **cosine similarity**
between each cell's raw detector signal and the AF spectrum it was assigned
to. A similarity of 1 means the cell's signal is perfectly explained by that
AF spectrum; lower values indicate a worse fit. Results are always reported
relative to a **mean-AF baseline** (every cell assigned to the mean AF
spectrum, regardless of which variant it actually resembles), so you can
immediately see whether your method beats the trivial case.

### Writing a compatible function

Your function must follow one of two calling conventions.

**Assign-type** functions take `(raw.data, spectra, af.spectra)` and return
an integer vector of per-cell AF-spectrum indices (one index per row of
`raw.data`, values in `1:nrow(af.spectra)`). These are the simpler of the two
to implement - you just need to decide which AF variant each cell is most
likely to be.

```{r assign-type-skeleton, eval=FALSE}
my.assign.af <- function( raw.data, spectra, af.spectra ) {
  # raw.data   : matrix, cells x detectors
  # spectra    : matrix, fluorophores x detectors (no AF row)
  # af.spectra : matrix, AF variants x detectors

  # ... your logic here ...

  af.idx  # integer vector, length nrow(raw.data), values in 1:nrow(af.spectra)
}
```

**Fit-type** functions (name must start with `"fit."`) receive pre-computed
OLS quantities and return both the assignments and the unmixed fluorophore
values. They are called with
`(raw.data, unmixed, unmixing.matrix, spectra, af.spectra)` and must return a
list with elements `$unmixed` (cells x fluorophores matrix, no AF column) and
`$af.idx` (integer vector as above).

```{r fit-type-skeleton, eval=FALSE}
fit.my.af <- function( raw.data, unmixed, unmixing.matrix, spectra, af.spectra ) {
  # unmixed         : matrix, cells x fluorophores (OLS result without AF)
  # unmixing.matrix : matrix, fluorophores x detectors

  # ... your logic here ...

  list(
    unmixed = unmixed_with_af_corrected,  # cells x fluorophores, no AF column
    af.idx  = af.idx                      # integer vector
  )
}
```

The naming convention matters: `test.af.accuracy()` dispatches on whether the
function name starts with `"fit."`, so make sure your function is named
accordingly and is available in the current search path (loaded via
`devtools::load_all()` or `source()`).

### Quick accuracy check: `test.af.accuracy()`

Use this first to confirm that your function actually works and to see how it
compares on a single unstained file.

```{r test-af-accuracy, eval=FALSE}
results <- test.af.accuracy(
  unstained.fcs = "path/to/unstained.fcs",
  spectra       = my.spectra,      # fluorophores x detectors
  af.spectra    = my.af.spectra,   # AF variants x detectors
  asp           = my.asp,
  functions     = c(
    "assign.af.fluorophores",  # existing method - keep as a reference point
    "assign.af.residuals",     # existing method
    "my.assign.af"             # your new function
  ),
  n.downsample  = 1000,        # events per run; increase for a final check
  plot.dir      = "figures/af_dev"
)
```

The function prints cosine similarity for each method and saves a PDF of
biplots - one panel per method plus a no-AF baseline - to `plot.dir`. You can
also inspect the numbers directly:

```{r inspect-results, eval=FALSE}
# mean cosine similarity for each method (higher is better)
sapply( results, `[[`, "Mean_Sim" )

# per-cell similarities for your method
hist( results[["my.assign.af"]]$Similarity, breaks = 50,
      main = "Per-cell cosine similarity", xlab = "Cosine similarity" )
```

### Scaling test: `benchmark.af.spectra.size()`

Once your function looks promising on a single file, use this to check whether
it holds up as the panel gets larger. It repeatedly subsamples the spectra
matrix to a range of panel sizes and runs `test.af.accuracy()` for each
subsample, producing a summary line plot of mean cosine similarity vs.
fluorophore count.

```{r benchmark-af-spectra-size, eval=FALSE}
bench <- benchmark.af.spectra.size(
  unstained.fcs = "path/to/unstained.fcs",
  spectra       = my.spectra,
  af.spectra    = my.af.spectra,
  asp           = my.asp,
  functions     = c(
    "assign.af.fluorophores",
    "assign.af.residuals",
    "my.assign.af"
  ),
  n.fluors     = c( 5, 10, 20, 30, 40 ),  # panel sizes to test
  n.draws      = 5,                        # random draws per panel size
  n.downsample = 1000,
  plot.dir     = "figures/af_dev"
)

# the summary data frame: mean and SD of cosine similarity across draws
head( bench$summary )

# the ggplot object, if you want to tweak the figure
bench$plot + ggplot2::ggtitle( "My new AF method" )
```

The output plot shows one line per method, with a ribbon spanning +/-1 SD across
random draws. A good new method should sit above the mean-AF baseline across
all panel sizes, and ideally narrow the gap on the existing methods or beat
them on larger panels where per-cell assignment becomes more important.

If you have results you'd like to share, feel free to open an issue or pull
request on GitHub with the summary data frame and the benchmark plot.

---

To install the `dev` branch:

```{r dev branch, eval=FALSE}
devtools::install_github("DrCytometer/AutoSpectral@dev")
```

To replace this with the (hopefully) stable version, run this:

```{r master branch, eval=FALSE}
remove.packages("AutoSpectral")
devtools::install_github("DrCytometer/AutoSpectral")
```



