---
title: "Development"
output: html_document
date: "2025-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here's some stuff that's in progress. These changes will show up first in the development `dev` branch, which you are welcome to install and test.

I don't have a particular timeframe or priority list for any of this, so ping me if any of this is important to you.

* Parallelization improvements. 
  + The parallelization of functions like `define.flow.control` can be problematic on Windows due to memory usage with `futures` and the tendency of Windows to usurp threads that R was using.
* More control over plotting, more color options. 
  + This should be rolled out in the current `dev` branch.
* A full workflow example with one of the datasets on Mendeley.
* A guide for the unmixing.
* More flexibility in negative/unstained samples.
* Allow multiple controls per fluorophore.
* An alternative to the automated gating.
* Better handling of file from the Northern Lights (3-laser Aurora).
* Better mapping of fluorophores to lasers in `spectral.trace`
* Some improvements/corrections to the weighting for WLS unmixing.
* Speed up per-cell fluorophore optimization.


To install the `dev` branch:

```{r dev branch, eval=FALSE}
devtools::install_github("DrCytometer/AutoSpectral@dev")
```

To replace this with the (hopefully) stable version, run this:

```{r master branch, eval=FALSE}
remove.packages("AutoSpectral")
devtools::install_github("DrCytometer/AutoSpectral")
```



